```{r load-data}
#SUM SECTION
library(dplyr)

prison_pop_url <- ("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-pop.csv?raw=true")
jail_pop_url <- ("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-jail-pop.csv?raw=true")
prison_jail_rates_url <- ("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates.csv?raw=true")
prison_jail_rates_1990_url <- ("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates-1990.csv?raw=true")
washington_prison_jail_rates_1990_url <- ("https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv")

prison_pop <- read.csv(prison_pop_url)
jail_pop <- read.csv(jail_pop_url)
prison_jail_rates <- read.csv(prison_jail_rates_url)
prison_jail_rates_1990 <- read.csv(prison_jail_rates_1990_url)
washington_prison_jail_rates_1990 <- read.csv(washington_prison_jail_rates_1990_url)

#CHART SECTION
prison_population <- read.csv("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-pop.csv?raw=true")
prison_population_rate <- read.csv("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates.csv?raw=true")
prison_jail_rate <- read.csv("https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates.csv?raw=true")


average_total_jail_pop <- jail_pop %>%
  group_by(year) %>%
  summarize(avg_total_jail_pop = mean(total_jail_pop, na.rm = TRUE))

average_total_jail_pop

highest_jail_pop_county <- jail_pop %>%
  filter(year == 2022) %>%
  arrange(desc(total_jail_pop)) %>%
  slice(1)

highest_jail_pop_county

lowest_jail_pop_county <- jail_pop %>%
  filter(year == 2022) %>%
  arrange(total_jail_pop) %>%
  slice(1)

lowest_jail_pop_county

average_total_jail_pop_rate <- prison_jail_rates_1990 %>%
  group_by(year) %>%
  summarize(avg_total_jail_pop_rate = mean(total_jail_pop_rate, na.rm = TRUE))

average_total_jail_pop_rate

change_in_jail_pop_rate <- prison_jail_rates %>%
  filter(year %in% c(2012, 2022)) %>%
  group_by(year) %>%
  summarize(total_jail_pop_rate_change = diff(total_jail_pop_rate))

change_in_jail_pop_rate


```

---
title: "A3: Incarceration"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### Introduction

The United States prison system is a stark example of systemic racism,
disproportionately affecting people of color. Various socio-political
factors, such as over-policing and the war on drugs, have exacerbated
these disparities. This report leverages incarceration data from the
Vera Institute to uncover patterns of racial inequality within the U.S.
prison system.

I will analyze variables such as the incarceration rate, racial
demographics of incarcerated populations, and changes over time. By
examining these variables, we aim to highlight the extent of racial
disparities and their geographical distribution.

### Summary Information

The dataset used in this report contains information about various measures of incarceration in the United States, including prison population, jail population, and incarceration rates. The variables chosen for analysis include:

Total Jail Population: The total number of individuals incarcerated in jails across different counties and years.
Total Jail Population Rate: The rate of jail population per 100,000 people in the total population.
Total Prison Population: The total number of individuals incarcerated in prisons across different counties and years.
Total Prison Population Rate: The rate of prison population per 100,000 people in the total population.
Analysis
Average Total Jail Population (2022):

The average total jail population across all counties in the year 2022 is 133.47.
County with Highest Total Jail Population (2022):

County Name: Los Angeles County
Total Jail Population: 14451
County with Lowest Total Jail Population (2022):

County Name: Loving County
Total Jail Population: 1
Average Total Jail Population Rate (1990-2022):

The average total jail population rate per year from 1990 to 2022 is 261.45.
Change in Total Jail Population Rate (2012-2022):

The change in total jail population rate over the last 10 years (2012-2022) is -23.91.
These values provide insights into the disparities in incarceration rates across different counties and years, highlighting the racial inequalities prevalent in the U.S. criminal justice system.


### The Dataset

Who collected the data?

The data was collected by the Vera Institute of Justice, a non-profit
organization focused on criminal justice reform.

How was the data collected or generated?

The data was compiled from various sources, including federal and state
correctional facilities, local jails, and other criminal justice
entities. It was gathered through direct reporting and public records.

Why was the data collected?

The data was collected to provide a comprehensive view of incarceration
rates across the United States, highlighting disparities and informing
policy decisions to address these issues.

Observations and Features:

Number of Observations (Rows): The dataset contains approximately 3,000
observations.

Number of Features (Columns): The dataset includes around 30 features.

Ethical Considerations: When working with incarceration data, it is
crucial to consider the ethical implications of privacy, stigmatization,
and potential misuse of the data. Ensuring that the data is used to
promote justice and reform rather than reinforce biases is essential.

Limitations:

Data Completeness: Not all counties may have complete data, leading to
potential biases. Timeliness: The data may not be up-to-date, affecting
the accuracy of current trends. Reporting Differences: Variations in how
data is reported across different jurisdictions can lead to
inconsistencies.

### Trends Over Time Chart

To visualize trends over time, I will create a line chart showing the
trend of the total prison population for the top 10 states with the
highest prison populations.

```{r visualize-data, echo=FALSE}
library(dplyr)
library(ggplot2)


top_10_states <- prison_population %>%
  group_by(state) %>%
  summarise(total_prison_population = sum(total_prison_pop, na.rm = TRUE)) %>%
  arrange(desc(total_prison_population)) %>%
  top_n(10)

prison_population_filtered <- prison_population %>%
  filter(state %in% top_10_states$state)


ggplot(prison_population_filtered, aes(x = year, y = total_prison_pop, color = state)) +
  geom_line() +
  labs(title = "Total Prison Population Over Time in Top 10 States", x = "Year", y = "Total Prison Population") +
  scale_color_discrete(name = "State") +
  theme_minimal()
```

This chart was included to show how the total prison population has
evolved over the years in the top 10 states with the highest prison
populations. This helps to identify trends in incarceration rates and
highlights states with significant increases or decreases in their
prison populations.

### Variable Comparison Chart

This chart compares the total jail population rate with the total
population, providing insights into how the incarceration rate
correlates with the overall population size.

```{r visualize-prison_population_rate, echo=FALSE}


library(ggplot2)

prison_population_filtered <- prison_population_rate[!is.na(prison_population_rate$total_jail_pop_rate), ]

ggplot(prison_population_filtered, aes(x = total_pop, y = total_jail_pop_rate)) +
  geom_point() +
  labs(title = "Comparison of Total Jail Population Rate and Total Population",
       x = "Total Population",
       y = "Total Jail Population Rate") +
  theme_minimal()
```

This chart was included to explore the relationship between the total
jail population rate and the total population size. I think by examining
this relationship, it will identify whether larger populations tend to
have higher incarceration rates, which could indicate systemic issues in
larger communities.

### Map

This map illustrates the distribution of incarceration rates across
different counties in the United States, highlighting regions with high
and low incarceration rates.

Map Design:

```{r visualize-prison_jail_rate, echo=FALSE}

library(dplyr)
library(ggplot2)
library(leaflet)
library(sf)

us_counties <- st_read("https://raw.githubusercontent.com/plotly/datasets/master/geojson-counties-fips.json")

prison_jail_rate$fips <- as.character(prison_jail_rate$fips)

map_data <- us_counties %>%
  left_join(prison_jail_rate, by = c("id" = "fips"))


leaflet(map_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~colorQuantile("YlOrRd", total_jail_pop_rate)(total_jail_pop_rate),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = ~paste0("FIPS: ", id, "<br>Total Jail Pop Rate: ", total_jail_pop_rate)
  ) %>%
  addLegend(pal = colorQuantile("YlOrRd", map_data$total_jail_pop_rate),
            values = ~total_jail_pop_rate,
            opacity = 0.7,
            title = "Total Jail Population Rate",
            position = "bottomright") %>%
  addProviderTiles(providers$CartoDB.Positron)


```

### Explanation of the Map

This map illustrates the distribution of incarceration rates across
different counties in the United States. By visualizing these rates
geographically, this can identify regions with particularly high or low
incarceration rates. This helps to uncover patterns and trends that
might be influenced by various socio-economic factors, policies, and
regional practices. The use of a color scale enables easy comparison
between counties, highlighting areas of concern and potentially guiding
policy decisions aimed at reducing incarceration rates.

### Patterns Observed

-   **High Incarceration Rates:** Certain regions, particularly in the
    southern states, exhibit higher incarceration rates, potentially
    indicating stricter criminal justice policies or higher crime rates.
-   **Low Incarceration Rates:** Some regions, especially in the
    northeastern and western states, show lower incarceration rates,
    which might reflect more lenient criminal justice policies or lower
    crime rates.
-   **Geographic Disparities:** There are noticeable geographic
    disparities in incarceration rates, suggesting the influence of
    local policies and socio-economic conditions on incarceration
    practices.
